The Evolution of Phenotypically Invariant Genetic Networks

### 2020-06-18 kalman_walk_paper.tex
added other refs

 % In biological systems, the relationship between structure and function is often not invertible -- that is information about function is not usually sufficient to determine structure.
 % Non-invertiblility thus entails a diversity of functionally equivalent structures.
  Biological systems, at nearly all levels of organization, display considerable degeneracy, or (perhaps more descriptively) a capacity for isofunctional \emph{structural plasticity} -- 
  that is systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}. 
  This is a consequence of the often non-invertibility (i.e. degeneracy) of the mapping between a system's structure and function.
Examples can be found across nearly all levels of biological organization from the level of the genetic code itself \citep{crick1963recent}
all the way up to the convergent evolution of adaptive traits. Between these levels, structural plasticity can be found in the \emph{neutral networks}
  of nucleic acid sequences that map to an RNA secondary structure \citep{gruner1996analysis}\revpoint{1}{5} and amino acid sequences that fold similarly \citep{babajide1997neutral},
as well as in the thermodynamic stability of proteins \citep{hart2014thermodynamic}. Further examples include the vast space of functionally equivalent
potential regulatory sequences \citep{hare2008sepsid}, 
in the logic of transcriptional \citep{tsong2006evolution, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription} 
and neural circuits \citep{trojanowski2014neural}, and in developmental systems \citep{true2001developmental}. 
\revpoint{1}{1} \revpoint{1}{3}
  \jss{Other ideas instead of "degeneracy" or "isofunctional structural plasticity": non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}

@@

Structural plasticity has been argued to be necessary for natural selection \citep{edelman2001degeneracy}, 
as it may function as a mechanism for biological robustness and evolvability, or manifest as 
\emph{canalization} \citep{whitacre2010degeneracy}. Degeneracy of the genetic code, and of genetic networks, 
for instance, can make sequences more fault-tolerant to mutations \citep{sonneborn1965degeneracy} and open up pathways to adaptation \citep{wagner2008robustness}.
Codon degeneracy can also enable the evolution of neutral and nearly neutral molecular changes \citep{kimura1968genetic},
and perhaps even contribute to the formation of new species \citep{gavrilets2014models}.

We are interested in the nature and evolutionary implications of structural plasticity and nonidentifiability in the context of gene regulatory networks.
Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
If systems architectures are not functionally unique, can this open up neutral evolutionary paths, and do
  these paths manifest as \emph{developmental system drift} \citep{true2001developmental} and contribute to speciation?
First, we apply results from system theory which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes.
Finally, we briefly outline and apply a quantitative genetics framework to estimate the speed at which system drift can lead to speciation via reproductive incompatibility,
which we develop in more detail in a companion paper \citep{schiffman2020genetic}.

>>>>

 % In biological systems, the relationship between structure and function is often not invertible -- that is information about function is not usually sufficient to determine structure.
 % Non-invertiblility thus entails a diversity of functionally equivalent structures.
  Biological systems, at nearly all levels of organization, display considerable degeneracy, or (perhaps more descriptively) a capacity for isofunctional \emph{structural plasticity} -- 
  that is systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}. 
  This is a consequence of the often non-invertibility (i.e. degeneracy) of the mapping between a system's structure and function.
Examples can be found across nearly all levels of biological organization from the level of the genetic code itself \citep{crick1963recent}
all the way up to the convergent evolution of adaptive traits. Between these levels, structural plasticity can be found in the \emph{neutral networks}
  of nucleic acid sequences that map to an RNA secondary structure \citep{gruner1996analysis}\revpoint{1}{5} and amino acid sequences that fold similarly \citep{babajide1997neutral},
as well as in the thermodynamic stability of proteins \citep{hart2014thermodynamic}. Further examples include the vast space of functionally equivalent
potential regulatory sequences \citep{hare2008sepsid}, 
in the logic of transcriptional \citep{tsong2006evolution, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription} 
and neural circuits \citep{trojanowski2014neural}, and in developmental systems \citep{true2001developmental}. 
\revpoint{1}{1} \revpoint{1}{3}
  \jss{Other ideas instead of ``degeneracy'' or ``isofunctional structural plasticity'': non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}

@@

Structural plasticity has been argued to be necessary for natural selection \citep{edelman2001degeneracy}, 
as it may function as a mechanism for biological robustness and evolvability, or manifest as 
\emph{canalization} \citep{whitacre2010degeneracy}. Structural plasticity of the genetic code, and of genetic networks, 
for instance, can make sequences more fault-tolerant to mutations \citep{sonneborn1965degeneracy} and open up pathways to adaptation \citep{wagner2008robustness}.\jss{Is it clear what ``structural plasticity of the genetic code'', as opposed to ``degeneracy of the genetic code'', means here?}
Structural plasticity can also enable the evolution of neutral and nearly neutral molecular changes \citep{kimura1968genetic},
and perhaps even contribute to the formation of new species \citep{gavrilets2014models}.

We are interested in the nature and evolutionary implications of structural plasticity and nonidentifiability in the context of gene regulatory networks.
Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
If systems architectures are not functionally unique, can this open up neutral evolutionary paths, and do
  these paths manifest as \emph{developmental system drift} \citep{true2001developmental} and contribute to speciation?
First, we apply results from system theory which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes.
Finally, we briefly outline and apply a quantitative genetics framework to estimate the speed at which system drift can lead to speciation via reproductive incompatibility.
In this approach, a population diffuses along the neutral ridges of a high-dimensional space of possible system parameters, in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}\revpoint{2}{12}. Our model is developed in more detail in a companion paper \citep{schiffman2020genetic}.
